---
title: "p8105_hw3_tc3326"
author: "KK Chen"
date: "2024-10-16"
output: github_document
---
# Problem 1

## write a short description of the dataset
```{r}
library(tidyverse)
library(p8105.datasets)

data("ny_noaa")

nrow(ny_noaa)
ncol(ny_noaa)

summary(ny_noaa)

na_summary <- colSums(is.na(ny_noaa))
na_summary
```
This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Key variables include weather station id, date of observation,  (tenths of mm), precipitation (tenths of mm), snowfall (mm), snow depth (mm), and min and max daily temperature (tenths of degrees C). The dataset contains some significant missing data. There are 145,838 missing values for precipitation, 381,221 for snowfall, and 591,786 for snow depth. 

## Data cleanning
```{r}
ny_noaa %>% 
  count(snow) %>%
  arrange(desc(n))

ny_noaa = 
  ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))
```
Comment:

0 mm snowfall is most common because snow is a seasonal occurrence. Most days throughout the year, particularly in non-winter months, do not see any snow.

Missing data (NA) for snowfall is the second most common. It may result from days or stations where snowfall was not measured.

25 mm, 13 mm, and 51 mm are the next most common values. Light snowfall events are more frequent than heavy snowstorms. Snowfall is often measured in the U.S. using inches, it's likely that these values were recorded as fractional inches and then converted to millimeters for consistency in this dataset. 

## Make a two-panel plot showing the average max temperature in January and in July in each station across years. 
```{r}
average_max_tem_plot <- ny_noaa %>%
  filter(month %in% c(1, 7)) %>%  
  group_by(id, year, month) %>%
  summarize(avg_tmax = mean(tmax, na.rm = TRUE), .groups = 'drop')  

ggplot(average_max_tem_plot, aes(x = year, y = avg_tmax, color = id)) +
  geom_line(alpha = 0.7) +  
  facet_wrap(~ month) + 
  labs(
    title = "Average Maximum Temperature in January and July In Each Station Across Years",
    x = "Year",
    y = "Average Max Temperature (°C)",
    color = "Station ID"
  ) +
  theme_minimal() +
  theme(legend.position = "none") 
```
Comments:

In both January and July, most stations show a fairly stable trend in maximum temperatures across the years, with some fluctuations. The temperatures in January are generally much lower than in July, consistent with winter and summer temperatures. Overall the trends remain within a similar range for most stations within a month across the years. 

Some extreme negative values (below -50°C) are present in the January panel.
There is one noticeable outlier on July of 1988 and a few other outliers in July panel.

## Make a two-panel plot showing tmax vs tmin for the full dataset 
```{r}
hex_plot1 <- ggplot(ny_noaa, aes(x = tmin, y = tmax)) +
  geom_hex(bins = 50) +  
  labs(
    title = "Hex Plot of Maximum Temperature vs Minimum Temperature",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)",
  ) +
  theme_minimal()
```

## Show the distribution of snowfall values greater than 0 and less than 100 separately by year
```{r}
snowfall_dist_plot2 <- ny_noaa %>%
  filter(snow > 0, snow < 100) %>%  
  ggplot(aes(x = factor(year), y = snow)) + 
  geom_boxplot() + 
  labs(
    title = "Distribution of Snowfall (0 < Snowfall < 100 mm) by Year",
    x = "Year",
    y = "Snowfall (mm)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  
```

```{r}
library(patchwork)
combined_plot12 <- hex_plot1 | snowfall_dist_plot2
```

```{r}
print (combined_plot12)
```
Comment:

The hex plot reveals that while most of the data points cluster tightly along a line where tmax is greater than tmin, there are some unusual cases where tmax is recorded as less than tmin. This is not physically possible and raises important concerns regarding data quality.

The boxplot for snowfall between 0 and 100 mm shows fairly consistent patterns across years, with the median snowfall amount remaining around 25 mm. This suggests that most snowfalls are relatively light or moderate.

The boxplots display some variability and outliers, particularly in years with snowfalls near the upper limit of the filter (close to 100 mm). These could represent unusually heavy snow events for those years.

# Problem 2

## Import and Organize data sets
```{r}
library(tidyverse)
```

```{r}
nhanes_accel <- read_csv("./data/nhanes_accel.csv") %>%
  janitor::clean_names()
```

```{r}
nhanes_covar <- read_csv("./data/nhanes_covar.csv", skip = 4)%>%
 set_names(c("seqn", "sex", "age", "BMI", "education")) %>%

mutate(
    seqn = as.integer(seqn),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    age = as.numeric(age),
    BMI = as.numeric(BMI),
    education = factor(education, levels = c(1, 2, 3), labels = c("Less than     high school", "High school equivalent", "More than high school"))
      ) %>%
  filter(age >= 21) %>%
  drop_na(sex, age, BMI, education)
```

```{r}
merged_data <- nhanes_accel %>%
  inner_join(nhanes_covar, by = "seqn")
```

## Produce a reader-friendly table for the number of men and women in each education category
```{r}
merged_data %>%
  group_by(sex, education) %>%
  summarize(count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = sex, values_from = count, values_fill = 0) %>%
  knitr::kable()
```
Comments: 
This table shows that both men and women in the dataset tend to have higher education levels, with the "More than high school" group having the highest representation for both genders. However, men are more prevalent in the "High school equivalent" category, while the "Less than high school" group shows an almost equal gender distribution. 

## Create a visualization of the age distributions for men and women in each education category
```{r}
ggplot(merged_data, aes(x = education, y = age, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Age Distributions by Sex and Education Level",
    x = "Education Level",
    y = "Age",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
Comments:
Participants with less education tend to be older, with the "less than high school" group having the oldest participants on average. In contrast, the "more than high school" group includes younger participants. 

In High School Equivalent,women in this group tend to be slightly older than men, with a median age closer to 60 for women and around 55 for men.Most women are of similar ages (closer to the median), while men show more variability.

In More than High school, it shows the most diverse age distribution for both men and women. It suggests that people with more than a high school education come from a broad range of age groups.The median age for both men and women is around 41 years, indicating that the central tendency is similar for both genders in this education level.

## Aggregate and create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis)

```{r}
activity_age_plot <- merged_data %>%
  mutate(total_activity = rowSums(across(min1:min1440), na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  facet_grid(.~ education) +
  labs(
    title = "Total Activity By Age and Education Level and Sex",
    x = "Age",
    y = "Total Activity (MIMS)",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(activity_age_plot)
```
Comment:
All groups show a decline in activity with age, participants with higher education maintain higher levels of activity into older age.
In the “Less than high school” group, physical activity decreases with age for both men and women, particularly after age 60.
In the “High school equivalent” group, middle age women generally have higher activity than middle age men. Activity decreases for women and men after age 40. 
In the “More than high school” group, activity trends are relatively stable across age, with men and women displaying similar activity patterns throughout the lifespan. However, women tend to have slightly higher activity levels than men overall. After the age of 60, both men and women experience a gradual decrease in total activity. The decline is more gradual compared to those with lower education levels.

## Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex 
```{r}
merged_data <- merged_data %>%
  pivot_longer(cols = min1:min1440, names_to = "minute", values_to = "activity") %>%
  mutate(minute = as.numeric(gsub("min", "", minute))) %>%
  group_by(education, sex, minute) %>%
  summarize(mean_act = mean(activity, na.rm = TRUE), .groups = 'drop')
```

```{r}
Hour_course_plot <- merged_data %>%
  ggplot(aes(x = minute, y = mean_act, color = sex)) +
  geom_line(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  facet_grid(.~ education) +
  labs(
    title = "24-Hour Activity Time Courses by Education Level and Sex",
    x = "Minute of the Day",
    y = "Mean Activity",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(Hour_course_plot)
```
Comments:

Across all education levels, activity peaks in the morning (around 8 a.m.) and declines in the evening. Women generally show higher activity levels, especially in the "High school equivalent" and "More than high school" groups.

Less than High School: Lower overall activity. There is a sharp peak in the morning and a steady decline throughout the day. Men show slightly higher activity than women.

High School Equivalent: Higher and more sustained activity levels than the Less than High School group. Women tend to have higher activity than men, especially during mid-day.

More than High School: This group has the highest activity levels throughout the day, with women consistently more active than men.

# Problem 3

## Import and clean datasets
```{r}
library(tidyverse)
library(janitor)
```

```{r}
combined_data <- read_csv("./data/Jan 2020 Citi.csv") %>%
  mutate(month = "jan", year = 2020) %>%
  clean_names() %>%
  bind_rows(
    read_csv("./data/July 2020 Citi.csv") %>%
      mutate(month = "jul", year = 2020) %>%
      clean_names(),
    read_csv("./data/Jan 2024 Citi.csv") %>%
      mutate(month = "jan", year = 2024) %>%
      clean_names(),
    read_csv("./data/July 2024 Citi.csv") %>%
      mutate(month = "jul", year = 2024) %>%
      clean_names()
  ) %>%
  distinct() 

obs <- combined_data %>%
  nrow()
summary(combined_data)
```
Describe the resulting dataset:

The dataset contains 99,485 Citi Bike rides from January and July of 2020 and 2024. Each ride has details such as ride ID, ride duration, bike type, the day of the week, start and end station names, and whether the rider is a member or casual user. 

## Produce a reader-friendly table 
```{r}
ride_summary <- combined_data %>%
  group_by(year, month, member_casual) %>%
  summarize(total_rides = n(), .groups = 'drop') %>%
  pivot_wider(names_from = member_casual, values_from = total_rides, values_fill = 0) %>%
    adorn_totals("row")
```

```{r}
print(ride_summary)
```
Comments:

The table shows that Citi Bike usage is significantly higher in July compared to January, reflecting strong seasonal variation. Member riders consistently account for the majority of rides, but casual ridership also sees a notable increase during the summer. Between 2020 and 2024, there is a marked growth in both casual and member rides. It indicates the increasing popularity of the system, especially during warmer months.

## Make a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations.

```{r}
top_stations_july_2024 <- combined_data %>%
  filter(month == "jul" & year == 2024) %>%
  group_by(start_station_name) %>%
  summarize(total_rides = n(), .groups = 'drop') %>%
  arrange(desc(total_rides)) %>%
  slice_head(n = 5)  
```

```{r}
top_stations_july_2024 %>%
  knitr::kable(col.names = c("Start Station", "Total Rides"), 
               caption = "Top 5 Most Popular Starting Stations for July 2024")
```

## Make a plot to investigate the effects of day of the week, month, and year on median ride duration. 
```{r}
median_duration_data <- combined_data %>%
  group_by(year, month, weekdays) %>%
  summarize(median_duration = median(duration, na.rm = TRUE), .groups = 'drop')

ggplot(median_duration_data, aes(x = weekdays, y = median_duration, fill = month)) +
  geom_bar(stat = "identity", position = "dodge") +  
  facet_wrap(~ year) +  
  labs(
    title = "Effect of Day of the Week, Month, and Year on Median Ride Duration",
    x = "Day of the Week",
    y = "Median Ride Duration (minutes)",
    fill = "Month"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```
Comments:

Across both 2020 and 2024 years, July consistently shows higher median ride durations compared to January. 

Weekend Days (Saturday and Sunday) tend to have the longest median ride durations, especially in July. Weekdays generally see shorter median ride durations.

2020 year shows a sharper contrast between weekend and weekday ride durations compared to 2024, especially in July, where weekend ride durations are significantly higher.

## For data in 2024, make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration.
```{r}
distribution_2024_plot <- combined_data %>%
  filter(year == 2024)

ggplot(distribution_2024_plot, aes(x = duration, fill = member_casual)) +
  geom_histogram(binwidth = 10, position = "dodge", alpha = 0.8) +  
  facet_grid(month ~ rideable_type) +  
  labs(
    title = "Ride Distribution of Ride Duration by Month, Membership Status, and Bike Type in 2024",
    x = "Ride Duration (minutes)",
    y = "Count",
    fill = "Member/Casual"
  ) +
  theme_minimal()
```
Comments:

Members have shorter ride durations compared to casual riders across both classic and electric bikes. This trend is visible in both January and July for electric bikes. Casual riders tend to have a wider spread in ride durations

Electric bikes show a concentration of rides with very short durations among members.Classic bikes tend to have a slightly more distributed ride duration pattern.

July sees more rides overall compared to January, which aligns with seasonal patterns where summer months attract more bike riders, both casual and member. Casual riders in July tend to have longer ride durations compared to January for classic bikes.

